Dependently Typed Multi-stage Calculus

### 2019-11-29 slides/main.tex
Fix.
Add related work.
The last work is adding the kinding judgement slide.

\documentclass[dvipdfmx,aspectratio=169, 20pt]{beamer}
\usepackage[absolute,overlay]{textpos}
\usepackage{amsmath}
\usepackage{amsthm}
\usepackage{ascmac}
\usepackage[backend=biber, style=numeric]{biblatex}
\usepackage{color}
\usepackage{mathtools}
\usepackage{url}
\usepackage{bcprules, proof}
\usepackage{xparse}
\usepackage{xspace}
\usepackage{fancybox}
\usepackage{float}
\usepackage{bcprules}
\usepackage{ebproof}
\usepackage{lscape}
\usepackage{pgfpages}
\usepackage{bm}

@@

\begin{frame}[fragile]{What is Multi-stage Programming}
    A programming paradigm that enables
    \begin{itemize}
        \item Generation of code at run-time
        \item Evaluation of code at run-time
    \end{itemize}
    Applications
    \begin{itemize}
        \item Speeding up programs using run-time information
    \end{itemize}
    \note{
    }
\end{frame}

\begin{frame}[fragile]{Example of Multi-stage Programming in MetaOCaml}
    \begin{exampleblock}{}
        \begin{verbatim}
 # let a = < (1 + 2) >     (* Make a code *)
 val a : int code = < (1 + 2) >
 # run a                   (* Run a code *)
 - : int = 3

 # let b = <(~a) + (~a)>   (* Embed codes *)
 val b : int code = < ((1 + 2) + (1 + 2)) >
 # run b
 - : int = 6
        \end{verbatim}
    \end{exampleblock}
    \note{
    }
\end{frame}

\begin{frame}[fragile]{Application of Multi-stage Programming: vadd-gen}
    Generate efficient vector addition functions for given length.
    \begin{exampleblock}{}
    \begin{verbatim}
 val vadd-gen : Int -> (Vec -> Vec -> Vec) code
 # let vadd3 = vadd-gen 3
 val vadd3 : (Vec -> Vec -> Vec) code
 (* vadd3 v w =                            *)
 (*   < let (v, w)=(~v, ~w) in             *)
 (*      [v[0]+w[0];v[1]+w[1];v[2]+w[2]] > *)
    \end{verbatim}
    \end{exampleblock}
    \note{
    }
\end{frame}

\begin{frame}[fragile]{The problem of vadd3 generated by vadd-gen}
    Runtime error occurs for vectors of wrong length.
    \begin{exampleblock}{}
    \begin{verbatim}
 val vadd3 : (Vec -> Vec -> Vec) code
 # (run vadd3) [1;2;3] [4;5;6]
 - : Vec = [5;7;9]
 # (run vadd3) [1;2] [4;5]
 Exception: Failure
    \end{verbatim}
    \end{exampleblock}
    \note{
    }
\end{frame}

\begin{frame}[fragile]{Dependent Types}
    \renewcommand{\V}{\text{Vec}\ }
    Types dependent on values
    \begin{itemize}
        \item Can prevent illegal use of {\verb|vadd3|}.
    \end{itemize}
    \begin{exampleblock}{}
 \begin{verbatim}
 [1;2]: Vec 2
 vadd3 : (Vec 3 -> Vec 3 -> Vec 3) code

 # (run vadd3) [1;2] [4;5]
 [1;2] has type Vec 2
 but was expected of type Vec 3
 \end{verbatim}
    \end{exampleblock}
    \note{
        By the way, we can write types dependent on values using dependent types.
    }
\end{frame}

\begin{frame}[fragile]{Our Approach: \LMD}
    Introduce dependent types into \LTP [Hanada\&Igarashi'14].
    \begin{itemize}
        \item Features of multi-stage programming
        \item Features of dependent types
    \end{itemize}
    Our contribution
    \begin{itemize}
        \item Prevent illegal use of generated code values with dependent types.
    \end{itemize}
    \note{
        Our approach to the problem of multi-stage programming is introducing dependent types into a multi-stage calculus by Hanada and Igarashi.
        We call our multi-stage calculus as \LMD.
    }
\end{frame}

@@

\begin{frame}[fragile]{Quick Overview of \LTP [Hanada\&Igarashi'14]}
    \begin{table}
        \begin{tabular}{ c | c | c }
            & \LTP & MetaOCaml \\[2mm]
            \hline
            Quote & \( \textbf{let } a = \TB (1 + 2) \) & \verb| let a = <(1+2)> | \\[2mm]
            Unquote & \( (\TB (\TBL a + \TBL a)) \) & \verb| <(~a)+(~a)> | \\[2mm]
            Run & \( \textbf{run } (\TB (1 + 2)) \) & \verb| run <(1+2)> | \\[2mm]
            Code type & \( \TW\ \text{Int} \) & \verb| int code |
        \end{tabular}
    \end{table}
\end{frame}

\begin{frame}[fragile]{Terms of \LTP}
    \begin{block}{Terms}
        \begin{tabbing}
            \hspace{5mm} \( M \) \= \( ::= c \) \hspace{20mm} \= (constants) \\
            \> \( \mid \TB M \) \> (quote) \\
            \> \( \mid \TBL M \) \> (unquote) \\
            \> \( \mid \textbf{run}\ M \) \> (run) \\
            \> \( \mid \% M \) \> (Cross Stage Persistence (CSP)) \\
            \> \( \mid x \mid \lambda x:\tau.M \mid M\ M \)
        \end{tabbing}
    \end{block}
    * Simplified version
\end{frame}

\begin{frame}[fragile]{Types of \LTP}
    \begin{block}{Types}
        \begin{tabbing}
            \hspace{5mm} \( \tau \) \= ::= X \hspace{20mm} \= (type-level constants) \\
            \> \( \mid \TW\ \tau \) \> (code type) \\
        \> \( \mid \tau \to \tau \) \> (function type)
    \end{tabbing}
    \end{block}

@@

    \begin{frame}[fragile]{What is a stage?}
        Level of nested quote
        \begin{exampleblock}{}
            \begin{center}
                \( \underbrace{(\textbf{run}\ (\TB\ \overbrace{(2 + 3)}^{\let\scriptstyle\textstyle\substack{\text{stage }1}})) + 1}_{\let\scriptstyle\textstyle\substack{\text{stage }0}} \)
            \end{center}
        \end{exampleblock}

    Typing judgements of \LTP include a stage.
    \begin{center}
        \( \G \vdash 2 + 3 : \I {\mathbf{@ 1}} \)
    \end{center}
    \note{
    }
\end{frame}

@@

\begin{frame}[fragile]{Cross Stage Persistence (CSP)}
    Enable using terms at lower stages at higher stages. \\[3mm]

    \begin{exampleblock}{}
    \begin{center}
        \( \textbf{let}\ \underbrace{\text{mul5}}_{\text{stage }0}\ x = 5 * x\ \textbf{in} \TB\ \underbrace{(\% \text{mul5}\ 4)}_{\text{stage }1} \)
    \end{center}
    \end{exampleblock}
    CSP symbol \( \% \) {\bf{lift up}} mul5 from stage 0 to stage 1.
\end{frame}

\section{Extension \LTP with Dependent Types}

@@

\begin{frame}[fragile]{Types and Kinds of \LMD}
    Append function dependent types to types of \LTP.
    \begin{block}{Types}
        \begin{tabbing}
            \hspace{5mm} \( \tau \) \= ::= X \hspace{20mm} \= (type-level constants) \\
            \> \( \mid \TW\ \tau \) \> (code type) \\
            \> \( \bm{\mid \Pi x:\tau.\tau} \) \> (function dependent type) \\
            \> \( \bm{\mid \tau\ M} \) \> (application, e.g. \(\text{Vec}\ 3\))
        \end{tabbing}
    \end{block}
    \begin{block}{Kinds}
        \( K ::= * \mid \Pi x:\tau.K \)
    \end{block}
    % Note: We need CSP to use normal variables in code types.
    % \begin{itemize}
    %     \item OK  \( \Pi n: \TW (\text{Vec}\ \%n) \)
    %     \item NG  \( \Pi n: \TW (\text{Vec}\ n) \)
    % \end{itemize}
    \note{
    }
\end{frame}

\begin{frame}[fragile]{Kinding rules of \LMD}
    Types have also stages because types may contains terms.
    \begin{center}
    {\footnotesize
    \infrule[{\KApp}]{
        \G\V \sigma:: (\Pi x:\tau.K) @ n \andalso \G\V M:\tau @ n
    }{
        \G\V \sigma\ M::K[x\mapsto M] @ n
    }
    \infrule[{\KAbs}]{
        \G\V \tau :: * @ n \andalso \G,x:\tau@ n \V \sigma::J@A
    }{
        \G\V(\Pi x:\tau.\sigma) :: (\Pi x:\tau.J) @ n
    }}
    \end{center}
    So, we need CSP in types also.
    \begin{exampleblock}{}
    \begin{center}
    \( \underbrace{\Pi n:\I. \TW (\overbrace{\text{Vec}\ \%n}^{\text{stage }1})}_{\text{stage }0} \)
    \end{center}
    \end{exampleblock}
\end{frame}

\begin{frame}[fragile]{Constants and type-level constants of \LMD}
    \begin{itemize}
        \item Constants: \( 0,1,2,3,\dots \)
        \item Type-level constants: Int, Vec, \dots
    \end{itemize}
    can be used at any stage.

    \begin{exampleblock}{}
        \begin{tabbing}
            \hspace{5mm} \= \text{val double-at-0} : \( \text{Vec }3 \to \text{Vec }3 \) \\[2mm]
            \> \textbf{let } \text{double} = \( \TB (\%\ \text{double-at-0}) \) \\
            \> \text{val double} : \( \TW \underbrace{(\text{Vec }3 \to \text{Vec }3)}_{\text{stage 1}} \)
        \end{tabbing}
    \end{exampleblock}
    Vec and integer 3 can be used at stage 1.
\end{frame}

\begin{frame}[fragile]{vadd-gen in \LMD}
    Type tells us the length of vector.
    \newcommand{\Vn}{\text{Vec}\ \%n}
    \newcommand{\Vt}{\text{Vec}\ \%3}
    \begin{exampleblock}{}
    \begin{tabbing}
        \hspace{5mm} \= \( \text{vadd-gen} : \Pi n:\I. \TW (\Vn \to \Vn \to \Vn) \) \\[2mm]
        \> \( \textbf{let}\ \text{vadd3} = \text{vadd-gen}\ 3;; \) \\
        \> \( \text{val}\ \text{vadd3} : \TW (\Vt \to \Vt \to \Vt) \) \\[2mm]
    \end{tabbing}
    \end{exampleblock}
    Now, vadd3 takes only vectors of length 3.
    \note{
        vadd-gen is changed as follows in \LMD.
        Thanks to dependent types ...
    }
\end{frame}

\begin{frame}[fragile]{Technical challenge of vadd3}
    Hard to compose with other functions.
    \newcommand{\Vn}{\text{Vec}\ \%n}
    \newcommand{\Vpt}{\text{Vec}\ \%3}
    \newcommand{\Vt}{\text{Vec}\ 3}
    \begin{exampleblock}{}
    \begin{tabbing}
        \hspace{5mm} \= \( \text{val}\ \text{vadd3} : \TW (\Vpt \to \Vpt \to \Vpt) \) \\
        \> \( \text{val}\ \text{double} : \TW (\Vt \to \Vt) \) \\[2mm]
        \> \( \textbf{let}\ \text{add-double}\ v\ w = \) \\
        \> \hspace{5mm} \( \TB (\underbrace{\TBL\text{double}}_{\Vt \to \Vt}\ \underbrace{(\TBL \text{vadd3}\ \TBL v\ \TBL w)}_{\Vpt});; \) \\
    \end{tabbing}
    \end{exampleblock}
    We cannot compose because of type mismatch.
\end{frame}

\begin{frame}[fragile]{Solution}
    \newcommand{\Vt}{\text{Vec}\ 3}
    \newcommand{\Vpt}{\text{Vec}\ \%3}
    Erase the CSP symbol \( \% \) under the condition \(M \) is closed.
    \begin{center}
        \infrule[{\QPercent}]{
            M \text{ is a closed term.}
        }{
            \G\vdash \% M \E M
        }
    \end{center}
    \begin{exampleblock}{}
    \begin{itemize}
        \item \( \text{Vec}\ \%3 \E \text{Vec}\ 3 \)
        \item \( \text{vadd3} : \TW (\Vt \to \Vt \to \Vt) \) \\
            \hspace{15mm} \( (\E  \TW (\Vpt \to \Vpt \to \Vpt)) \)
    \end{itemize}
    \end{exampleblock}
\end{frame}

@@

\begin{frame}[fragile]{Related Work}
    \begin{itemize}
        \item \LTP [Hanada\&Igarashi'14]
            \begin{itemize}
                \item Multi-stage programming including explicit CSP
            \end{itemize}
        \item Concoqtion [Forgarty et al.'14]
            \begin{itemize}
                \item Introduce indexed type into MetaOCaml.
                \item Types are indexed by Coq terms.
            \end{itemize}
    \end{itemize}
    \note{
    }
\end{frame}

\begin{frame}[fragile]{Conclusion}
    We introduced dependent types into multi-stage calculus.
    \begin{itemize}
        \item Avoid illegal use of generated code values with dependent types.
        \item Prove type soundness.
        \item Future work: algorithmic typing
    \end{itemize}
    \note{
    }
\end{frame}

@@

% \begin{frame}[fragile]{Generated functions by vadd-gen in \LMD}
%     Type system prevents illegal use of generated function.
%     \renewcommand{\V}{\text{Vec}}
%     \begin{tabbing}
%         \( \text{vadd3}\ [1;2;3]\ [4;5;6];; \) \\
%         \( - : \V = [5;7;9] \) \\[2mm]
%         \( \text{vadd3}\ [1;2;3;4]\ [4;5;6;7];; \) \\
%         \( \longrightarrow \textbf{TYPE ERROR!} \text{ (not RUNTIME ERROR)} \)
%     \end{tabbing}
%     \note{
%         Type of generated functions by vadd-gen in \LMD have the length information in their types.
%     }
% \end{frame}

@@

% \begin{frame}[fragile]{Types of \LMD}
%     \begin{block}{Types}
%     \( \tau,\sigma ::= X \mid \Pi x:\tau.\sigma \mid \tau\ M \mid \TW_{\alpha} \tau \mid \F\alpha.\tau \)
%     \end{block}
%     \begin{itemize}
%         \item \( X \): type-level constants 
%             \begin{itemize}
%                 \item \( \text{Vec} \), \( \text{Int} \).
%             \end{itemize}
%         \item \( \Pi x:\tau.\sigma \): dependent type.
%         \item \( \tau\ M \): Application of term to dependent type.
%             \begin{itemize}
%                 \item \( \text{Vec } 3 \)
%             \end{itemize}
%         \item \( \TW_{\alpha} \tau \): a type of code of type \( \tau \).
%             \begin{itemize}
%                 \item \( \text{vadd-gen 3} : \TW_\alpha (\text{Vec}\ 3 \to \text{Vec}\ 3 \to \text{Vec}\ 3) \)
%             \end{itemize}
%     \end{itemize}
%     \note{
%     }
% \end{frame}

% \begin{frame}[fragile]{Kinds of \LMD}
%     \begin{block}{Kinds}
%         \( K ::= * \mid \Pi x:\tau.K \)
%     \end{block}
%     \begin{itemize}
%         \item \( * \): kind of proper type
%             \begin{itemize}
%                 \item \( \text{Int} :: * \)
%             \end{itemize}
%         \item \( \Pi x:\tau.K \): kind dependent on term.
%             \begin{itemize}
%                 \item \( \text{Vec} :: (\Pi x:\text{Int}.*) \)
%             \end{itemize}
%     \end{itemize}
%     \note{
%     }
% \end{frame}

@@

% \begin{frame}[fragile]{What is signature \( \Sigma \)\ ?}
%     Signature \( \Sigma \) is a sequence of type-level constants and constant values.
%     \begin{itemize}
%         \item For example,
%             \begin{itemize}
%                 \item \( \Sigma = \text{Int}::*, \text{Vec}::\Pi x:\text{Int}.*, 0:\text{Int} \)
%             \end{itemize}
%         \item Elements of \( \Sigma \) can be used at any stages.
%     \end{itemize}
%     \note{
%         自然数が無限個あるというツッコミを誘発する気がする。
%         後者関数が存在することにするか?
%     }
% \end{frame}

@@

% \begin{frame}[fragile]{Example of Kinding Derivation}
%     \renewcommand{\Vec}{\text{Vec}}
%     \( \V \TB_\alpha [1,2,3] : \underbrace{\TW_\alpha (\Vec\ 3)}_{\TW_\alpha (\Vec\ 3) :: * @ \varepsilon} @ \varepsilon \) \\[2mm]
%     \pause
%     \( \Sigma = \Vec::\Pi x:\I.*, 0:\I, 1:\I, \cdots \)
%     \begin{center}
%         \footnotesize
%         \begin{minipage}{0.7\hsize}
%             \infer[\KTW]
%             {\V \TW_\alpha (\Vec\ 3) :: * @ \varepsilon}
%             {
%                 \infer[\KApp]
%                 {\V \Vec\ 3 :: * @ \alpha}
%                 {
%                     \infer[\KTConst]
%                     {\V \Vec :: \Pi x:\I.* @ \alpha}
%                     {\Vec :: \Pi x:\I.* \in \Sigma}
%                     \andalso
%                     \infer[\TConst]
%                     {\V 3 : \I @ \alpha}
%                     {3 : \I \in \Sigma}
%                 }
%             }
%         \end{minipage}
%     \end{center}
%     We can use \Vec\ and integers at any stages.
%     \note{
%         The constructed type Vec 3 is stage sensitive so we cannot use at arbitrary stages
%         but the elements, Vec and 3 are in the signature \( \Sigma \) and can be used at any stages.
%     }
% \end{frame}

>>>>

\documentclass[dvipdfmx,aspectratio=169, 20pt]{beamer}
\usepackage[absolute,overlay]{textpos}
\usepackage{amsmath}
\usepackage{amsthm}
\usepackage{ascmac}
\usepackage[backend=biber, style=numeric]{biblatex}
\usepackage{color}
\usepackage{mathtools}
\usepackage{url}
\usepackage{bcprules, proof}
\usepackage{xparse}
\usepackage{xspace}
\usepackage{fancybox}
\usepackage{fancyvrb}
\usepackage{float}
\usepackage{bcprules}
\usepackage{ebproof}
\usepackage{lscape}
\usepackage{pgfpages}
\usepackage{bm}

@@

\begin{frame}[fragile]{What is Multi-stage Programming}
    A programming paradigm that enables
    \begin{itemize}
        \item Generation of code at run-time
        \item Evaluation of code at run-time
    \end{itemize}
    Applications
    \begin{itemize}
        \item Speeding up programs using run-time information \\
            (a.k.a. run-time specialization)
        \item DSLs
    \end{itemize}
    \note{
        a.k.a = Also known as
    }
\end{frame}

\begin{frame}[fragile]{Example of Multi-stage Programming in MetaOCaml}
    \begin{exampleblock}{}
        \begin{Verbatim}[numbers=left]
 # let a = < (1 + 2) >     (* Make code *)
 val a : int code = < (1 + 2) >
 # run a                   (* Run code *)
 - : int = 3

 # let b = <(~a) + (~a)>   (* Embed code *)
 val b : int code = < ((1 + 2) + (1 + 2)) >
 # run b
 - : int = 6
        \end{Verbatim}
    \end{exampleblock}
    \note{
    }
\end{frame}

\begin{frame}[fragile]{Application of Multi-stage Programming: vadd-gen}
    Generate efficient vector addition functions for given length.
    \begin{exampleblock}{}
    \begin{Verbatim}[numbers=left]
 val vadd-gen : int -> (vec -> vec -> vec) code
 # let vadd3 = vadd-gen 3
 val vadd3 : (vec -> vec -> vec) code
 (* vadd3 = < fun v w ->                 *)
 (*   let (v, w)=(~v, ~w) in             *)
 (*   [v[0]+w[0];v[1]+w[1];v[2]+w[2]] >  *)
    \end{Verbatim}
    \end{exampleblock}
    \note{
    }
\end{frame}

\begin{frame}[fragile]{The problem of vadd3 generated by vadd-gen}
    Runtime error occurs for vectors of wrong length.
    \begin{exampleblock}{}
    \begin{Verbatim}[numbers=left]
 val vadd3 : (vec -> vec -> vec) code
 # (run vadd3) [1;2;3] [4;5;6]
 - : vec = [5;7;9]
 # (run vadd3) [1;2] [4;5]
 Exception: Failure
    \end{Verbatim}
    \end{exampleblock}
    \note{
    }
\end{frame}

\begin{frame}[fragile]{Our Approach: Dependent Types [Martin-L{\"o}f'73]}
    \renewcommand{\V}{\text{vec}\ }
    Types dependent on values
    \begin{itemize}
        \item Can prevent illegal use of {\verb|vadd3|}.
    \end{itemize}
    \begin{exampleblock}{}
 \begin{Verbatim}[numbers=left]
 [1;2]: vec 2
 vadd3 : (vec 3 -> vec 3 -> vec 3) code

 # (run vadd3) [1;2] [4;5]
 [1;2] has type vec 2
 but was expected of type vec 3
 \end{Verbatim}
    \end{exampleblock}
    \note{
        By the way, we can write types dependent on values using dependent types.
    }
\end{frame}

\begin{frame}[fragile]{Our Contribution: \LMD}
    We design a multi-stage calculus \LMD by
    introducing dependent types into \LTP [Hanada\&Igarashi'14].
    \begin{itemize}
        \item Features of multi-stage programming
        \item Features of dependent types
    \end{itemize}
    Technically
    \begin{itemize}
            \item Define well-formed kind, kinding, typing and equivalence rules. 
                \item Prove type soundness.
    \end{itemize}
    \note{
        Our approach to the problem of multi-stage programming is introducing dependent types into a multi-stage calculus by Hanada and Igarashi.
        We call our multi-stage calculus as \LMD.
    }
\end{frame}

@@

\begin{frame}[fragile]{Quick Overview of \LTP [Hanada\&Igarashi'14]}
    \begin{table}
        \begin{tabular}{ c | c | c }
            & \LTP & MetaOCaml \\[2mm]
            \hline
            Quote & \( \textbf{let } a = \TB (1 + 2) \) & \verb| let a = <(1+2)> | \\[2mm]
            Unquote & \( (\TB (\TBL a + \TBL a)) \) & \verb| <(~a)+(~a)> | \\[2mm]
            Code type & \( \TW\ \text{int} \) & \verb| int code |
        \end{tabular}
    \end{table}
\end{frame}

\begin{frame}[fragile]{Terms of \LTP}
    \begin{block}{Terms}
        \begin{tabbing}
            \hspace{5mm} \( M \) \= \( ::= c \) \hspace{20mm} \= (constants) \\
            \> \( \mid \TB M \) \> (quote) \\
            \> \( \mid \TBL M \) \> (unquote) \\
            \> \( \mid \% M \) \> (Cross Stage Persistence (CSP)) \\
            \> \( \mid x \mid \lambda x:\tau.M \mid M\ M \)
        \end{tabbing}
    \end{block}
    * Simplified version
\end{frame}

\begin{frame}[fragile]{Types of \LTP}
    \begin{block}{Types}
        \begin{tabbing}
            \hspace{5mm} \( \tau \) \= ::= X \hspace{20mm} \= (type-level constants, e.g., int, vec) \\
            \> \( \mid \TW\ \tau \) \> (code type) \\
        \> \( \mid \tau \to \tau \) \> (function type)
    \end{tabbing}
    \end{block}

@@

    \begin{frame}[fragile]{What is a stage?}
        Level of nested quote
        \begin{exampleblock}{}
            \begin{center}
                \( \underbrace{\TB\ \overbrace{(\lambda x:\text{int}.x + 1)}^{\let\scriptstyle\textstyle\substack{\text{stage }1}}}_{\let\scriptstyle\textstyle\substack{\text{stage }0}} \)
            \end{center}
        \end{exampleblock}

    Typing judgements and environments of \LTP include a stage.
    \begin{center}
        \( x:\text{int}@1 \vdash x + 1 : \text{int} {\mathbf{@ 1}} \)
    \end{center}
    \note{
    }
\end{frame}

@@

\begin{frame}[fragile]{Cross Stage Persistence (CSP)}
    Allow terms at lower stages to be used at higher stages. \\[3mm]

    \begin{exampleblock}{}
    \begin{center}
        \( \textbf{let}\ \underbrace{\text{mul5}}_{\text{stage }0}\ x = 5 * x\ \textbf{in} \TB\ \underbrace{(\% \text{mul5}\ 4)}_{\text{stage }1} \)
    \end{center}
    \end{exampleblock}
    CSP symbol \( \% \) {\bf{lifts up}} mul5 from stage 0 to stage 1. \\
    (In MetaOCaml, variables are lifted up implicitly.)
\end{frame}

\section{\LMD: Extension of \LTP with Dependent Types}

@@

\begin{frame}[fragile]{Overview of \LMD}
    Extension of multi-stage calculus \LTP with dependent types
    \begin{itemize}
        \item \LMD terms are the same as \LTP.
        \item Types are extended with dependent function type.
        \item There are equivalence rules for dependent types.
    \end{itemize}
\end{frame}

\begin{frame}[fragile]{Types and Kinds of \LMD}
    Append dependent function types to types of \LTP.
    \begin{block}{Types}
        \begin{tabbing}
            \hspace{5mm} \( \tau \) \= ::= X \hspace{20mm} \= (type-level constants) \\
            \> \( \mid \TW\ \tau \) \> (code type) \\
            \> \( \bm{\mid \Pi x:\tau.\tau} \) \> (dependent function type) \\
            \> \( \bm{\mid \tau\ M} \) \> (application, e.g. \(\text{vec}\ 3\))
        \end{tabbing}
    \end{block}
    \begin{block}{Kinds}
        \( K ::= * \mid \Pi x:\tau.K \)
    \end{block}
    % Note: We need CSP to use normal variables in code types.
    % \begin{itemize}
    %     \item OK  \( \Pi n: \TW (\text{vec}\ \%n) \)
    %     \item NG  \( \Pi n: \TW (\text{vec}\ n) \)
    % \end{itemize}
    \note{
    }
\end{frame}

\begin{frame}[fragile]{Kinding rules of \LMD}
    Types have also stages because types may contains terms.
    \begin{center}
    {\footnotesize
    \infrule[{\KApp}]{
        \G\V \sigma:: (\Pi x:\tau.K) @ n \andalso \G\V M:\tau @ n
    }{
        \G\V \sigma\ M::K[x\mapsto M] @ n
    }
    \infrule[{\KAbs}]{
        \G\V \tau :: * @ n \andalso \G,x:\tau@ n \V \sigma::J@A
    }{
        \G\V(\Pi x:\tau.\sigma) :: (\Pi x:\tau.J) @ n
    }}
    \end{center}
    So, we need CSP in types also.
    \begin{exampleblock}{}
    \begin{center}
    \( \underbrace{\Pi n:\text{int}. \TW (\overbrace{\text{vec}\ \%n}^{\text{stage }1})}_{\text{stage }0} \)
    \end{center}
    \end{exampleblock}
\end{frame}

\begin{frame}[fragile]{Constants and type-level constants of \LMD}
    \begin{itemize}
        \item Constants: \( 0,1,2,3,\dots \)
        \item Type-level constants: int, vec, \dots
    \end{itemize}
    can be used at any stage.
    \begin{exampleblock}{}
        \begin{tabbing}
            \hspace{5mm} \= \text{val double-at-0} : \( \text{vec }3 \to \text{vec }3 \) \\[2mm]
            \> \textbf{let } \text{double} = \( \TB (\%\ \text{double-at-0}) \) \\
            \> \text{val double} : \( \TW \underbrace{(\text{vec }3 \to \text{vec }3)}_{\text{stage 1}} \)
        \end{tabbing}
    \end{exampleblock}
\end{frame}

\begin{frame}[fragile]{vadd-gen in \LMD}
    Type tells us the length of vector.
    \newcommand{\Vn}{\text{vec}\ \%n}
    \newcommand{\Vt}{\text{vec}\ \%3}
    \begin{exampleblock}{}
    \begin{tabbing}
        \hspace{5mm} \= \( \text{vadd-gen} : \Pi n:\text{int}. \TW (\Vn \to \Vn \to \Vn) \) \\[2mm]
        \> \( \textbf{let}\ \text{vadd3} = \text{vadd-gen}\ 3;; \) \\
        \> \( \text{val}\ \text{vadd3} : \TW (\Vt \to \Vt \to \Vt) \) \\[2mm]
    \end{tabbing}
    \end{exampleblock}
    Now, vadd3 takes only vectors of length 3.
    \note{
        vadd-gen is changed as follows in \LMD.
        Thanks to dependent types ...
    }
\end{frame}

\begin{frame}[fragile]{Technical challenge of vadd3}
    Hard to compose with other functions.
    \newcommand{\Vn}{\text{vec}\ \%n}
    \newcommand{\Vpt}{\text{vec}\ \%3}
    \newcommand{\Vt}{\text{vec}\ 3}
    \begin{exampleblock}{}
    \begin{tabbing}
        \hspace{5mm} \= \( \text{val}\ \text{vadd3} : \TW (\Vpt \to \Vpt \to \Vpt) \) \\
        \> \( \text{val}\ \text{double} : \TW (\Vt \to \Vt) \) \\[2mm]
        \> \( \textbf{let}\ \text{add-double}\ v\ w = \) \\
        \> \hspace{5mm} \( \TB (\underbrace{\TBL\text{double}}_{\Vt \to \Vt}\ \underbrace{(\TBL \text{vadd3}\ \TBL v\ \TBL w)}_{\Vpt});; \) \\
    \end{tabbing}
    \end{exampleblock}
    We cannot compose because of type mismatch.
\end{frame}

\begin{frame}[fragile]{Solution}
    \newcommand{\Vt}{\text{vec}\ 3}
    \newcommand{\Vpt}{\text{vec}\ \%3}
    Erase the CSP symbol \( \% \) when \(M \) is closed (\QPercent).
    {\footnotesize{
    \begin{center}
        \infrule[{\QPercent}]{
            M \text{ is a closed term.}
        }{
            \G\vdash \% M \E M
        } \\[2mm]
        \infrule[{\TConv}]{
            \G\V M:\tau@n \andalso
            \G\V \tau\equiv \sigma @n
        }{
            \G\V M:\sigma@n
        }
    \end{center}
    }}
    \begin{exampleblock}{}
        \begin{itemize}
            \item \( \text{vec}\ \%3 \E \text{vec}\ 3 \)
            \item \( \text{vadd3} : \TW (\Vt \to \Vt \to \Vt) \) \\
                \hspace{15mm} \( (\E  \TW (\Vpt \to \Vpt \to \Vpt)) \)
        \end{itemize}
    \end{exampleblock}
\end{frame}

@@

\begin{frame}[fragile]{Related Work}
    \begin{itemize}
        \item \LTP [Hanada\&Igarashi'14]
        \item Concoqtion [Forgarty et al.'07]
            \begin{itemize}
                \item Introduce indexed type into MetaOCaml.
                \item Types are indexed by Coq terms. \\ (not full dependent types)
            \end{itemize}
        \item \( \lambda_{H\circ} \) [Pasalic'04]
            \begin{itemize}
                \item Meta-programming with dependent types
                \item No run and CSP
            \end{itemize}
    \end{itemize}
    \note{
    }
\end{frame}

\begin{frame}[fragile]{Conclusion}
    We introduced dependent types into multi-stage calculus.
    \begin{itemize}
        \item Avoid illegal use of generated code values with dependent types.
            \begin{itemize}
                \item Full dependent types not restricted ones.
            \end{itemize}
        \item Define equivalence rules.
            \begin{itemize}
                \item Especially, for CSP
            \end{itemize}
        \item Prove type soundness.
        \item Future work: algorithmic typing
    \end{itemize}
    \note{
    }
\end{frame}

@@

% \begin{frame}[fragile]{Generated functions by vadd-gen in \LMD}
%     Type system prevents illegal use of generated function.
%     \renewcommand{\V}{\text{vec}}
%     \begin{tabbing}
%         \( \text{vadd3}\ [1;2;3]\ [4;5;6];; \) \\
%         \( - : \V = [5;7;9] \) \\[2mm]
%         \( \text{vadd3}\ [1;2;3;4]\ [4;5;6;7];; \) \\
%         \( \longrightarrow \textbf{TYPE ERROR!} \text{ (not RUNTIME ERROR)} \)
%     \end{tabbing}
%     \note{
%         Type of generated functions by vadd-gen in \LMD have the length information in their types.
%     }
% \end{frame}

@@

% \begin{frame}[fragile]{Types of \LMD}
%     \begin{block}{Types}
%     \( \tau,\sigma ::= X \mid \Pi x:\tau.\sigma \mid \tau\ M \mid \TW_{\alpha} \tau \mid \F\alpha.\tau \)
%     \end{block}
%     \begin{itemize}
%         \item \( X \): type-level constants 
%             \begin{itemize}
%                 \item \( \text{vec} \), \( \text{int} \).
%             \end{itemize}
%         \item \( \Pi x:\tau.\sigma \): dependent type.
%         \item \( \tau\ M \): Application of term to dependent type.
%             \begin{itemize}
%                 \item \( \text{vec } 3 \)
%             \end{itemize}
%         \item \( \TW_{\alpha} \tau \): a type of code of type \( \tau \).
%             \begin{itemize}
%                 \item \( \text{vadd-gen 3} : \TW_\alpha (\text{vec}\ 3 \to \text{vec}\ 3 \to \text{vec}\ 3) \)
%             \end{itemize}
%     \end{itemize}
%     \note{
%     }
% \end{frame}

% \begin{frame}[fragile]{Kinds of \LMD}
%     \begin{block}{Kinds}
%         \( K ::= * \mid \Pi x:\tau.K \)
%     \end{block}
%     \begin{itemize}
%         \item \( * \): kind of proper type
%             \begin{itemize}
%                 \item \( \text{int} :: * \)
%             \end{itemize}
%         \item \( \Pi x:\tau.K \): kind dependent on term.
%             \begin{itemize}
%                 \item \( \text{vec} :: (\Pi x:\text{int}.*) \)
%             \end{itemize}
%     \end{itemize}
%     \note{
%     }
% \end{frame}

@@

% \begin{frame}[fragile]{What is signature \( \Sigma \)\ ?}
%     Signature \( \Sigma \) is a sequence of type-level constants and constant values.
%     \begin{itemize}
%         \item For example,
%             \begin{itemize}
%                 \item \( \Sigma = \text{int}::*, \text{vec}::\Pi x:\text{int}.*, 0:\text{int} \)
%             \end{itemize}
%         \item Elements of \( \Sigma \) can be used at any stages.
%     \end{itemize}
%     \note{
%         自然数が無限個あるというツッコミを誘発する気がする。
%         後者関数が存在することにするか?
%     }
% \end{frame}

@@

% \begin{frame}[fragile]{Example of Kinding Derivation}
%     \renewcommand{\vec}{\text{vec}}
%     \( \V \TB_\alpha [1,2,3] : \underbrace{\TW_\alpha (\vec\ 3)}_{\TW_\alpha (\vec\ 3) :: * @ \varepsilon} @ \varepsilon \) \\[2mm]
%     \pause
%     \( \Sigma = \vec::\Pi x:\text{int}.*, 0:\text{int}, 1:\text{int}, \cdots \)
%     \begin{center}
%         \footnotesize
%         \begin{minipage}{0.7\hsize}
%             \infer[\KTW]
%             {\V \TW_\alpha (\vec\ 3) :: * @ \varepsilon}
%             {
%                 \infer[\KApp]
%                 {\V \vec\ 3 :: * @ \alpha}
%                 {
%                     \infer[\KTConst]
%                     {\V \vec :: \Pi x:\text{int}.* @ \alpha}
%                     {\vec :: \Pi x:\text{int}.* \in \Sigma}
%                     \andalso
%                     \infer[\TConst]
%                     {\V 3 : \text{int} @ \alpha}
%                     {3 : \text{int} \in \Sigma}
%                 }
%             }
%         \end{minipage}
%     \end{center}
%     We can use \vec\ and integers at any stages.
%     \note{
%         The constructed type vec 3 is stage sensitive so we cannot use at arbitrary stages
%         but the elements, vec and 3 are in the signature \( \Sigma \) and can be used at any stages.
%     }
% \end{frame}
